The Concurrent Calculi Formalisation Benchmark

### 2024-02-09 short-paper/short-paper.tex
Rearrange related work sections.

% - Introduction
%   - POPLMark accelerated progress in mechanisation tools for PL
%     - POPLMark Reloaded tried to do the same for logical relations
%     - But they were deliberately limited in scope to not include process calculi
%   - We propose a new collection of benchmark problems for process calculi
%     - Why not concurrency in general? (after this point, only mention process calculi)
%     - Why is there a problem with process calculus mechanisations?
%     - Our experience says there is, and we will cite other experiences below
%   - We have noticed three key issues
%     - They are not covered by POPLMark
%     - There are fewer techniques than for binders
%     - Others also mention them as problems: cite and explain references briefly for each challenge topic
%   - Our goals are to determine the state of the art, develop best practices and tutorials, and identify improvements to proof assistant technology that would materially help the situation
\section{Introduction}
The POPLMark challenge~\cite{POPLMark} accelerated the development of proof assistants, libraries, and best practices for programming language mechanisation by establishing a common ground for systematically evaluating mechanisation techniques.
In doing so, it spearheaded a shift in the programming language research community towards publications that include mechanised proofs.
POPLMark Reloaded~\cite{POPLMarkReloaded} proposed a similar program for proofs involving logical relations.
These challenges, and therefore the developments they spurred, were deliberately limited in scope, but their authors noted they need for future benchmarks to tackle such areas as coinduction and linear environments.

@@

\todo{These paragraphs are still quite messy and unfocused.}
\paragraph{Linearity.}
In mechanised meta-theory, addressing linearity means choosing an appropriate representation of the linear context.
The crux of the matter is \emph{context splitting}, as showcased by the typing rule for parallel composition.
While a linear context is perhaps best seen as a multi-set, most proof assistants have better support for list processing.
The latter representation is intuitive, but may require establishing a large number of technical lemmata that are orthogonal to the mathematics of the problem under study, say proving type preservation for session types.
In this space, several designs are possible: one can label occurrences of resources to constrain their usage (\eg~\cite{CicconeP20}) or \alb{something --- ask Francisco} as in \cite{Castro2020}.
Some authors instead impose a multiset structure over lists (\eg~\cite{ChaudhuriLR19,Danielsson12}); still, the effort required to develop the infrastructure is significant.

@@

% Start of related work
\cite{Maksimovic2015} does not have name restriction, thus dodging the challenge of scope extrusion.

@@

Coinduction is widely supported in proof assistants.
Isabelle, Coq, and Abella seamlessly support coinductive definitions and reasoning.
Agda and Beluga support it through the convenient abstraction of co-patterns~\cite{Abel2013}.
Newer entrants to the field like the Lean proof assistant are working on adding a notion of coinduction to their system~\cite{Avigad2019,Keizer2023}.

@@

@@

% - Design of the benchmark
%   - We first list some general design remarks, then for each challenge
%   - General remarks
%     - We do not aim to be comprehensive (list other topics)
%     - However, we believe the best practices developed could also be of use in those topics
%     - We base our challenges on basic textbook theory with interesting proof techniques
%     - The challenges are independent (maybe note ``failure'' of POPLMark that everyone only did the first part)
%     - The challenges are as minimal as possible
%   - Linearity
%     - Type preservation for session types
%     - The main issue is context splitting
%     - Session types are pervasive in the field and simple to understand
%     - Binding two names makes duality obvious
%     - Brief overview of approaches and issues with them
%   - Scope extrusion
%     - Process calculi have several binders, but only restriction is different from ``normal'' ones
%     - The main issue is scope extrusion
%     - The pros and cons of established techniques for handling binders are different for process calculi, and it is unclear how exactly
%     - Brief overview of approaches and issues with them
%   - Coinduction
%     - Strong barbed bisimulation and congruence
%     - The main issue is the encoding of coinductive up-to techniques
%     - Strong equivalences to simplify (we don't have to abstract over internal transitions)
%     - Why barbed equivalences?
%     - No delegation to simplify
%     - Brief overview of approaches and issues with them
%   - Evaluation criteria
%     - We care not only about solutions, but about experiences and comparisons
%     - Mechanisation overhead
%     - Adequacy of the formal statements
%     - Cost of entry for tools and techniques
%     - Slight changes are OK if they are justified and adequate
\section{Design of the Benchmark}\label{sec:design-discussion}
\todo{Add references to support our claims and decisions in this whole section.}

@@

\subsubsection{Linearity.}
For our challenge on linear reasoning we have chosen a type preservation theorem
for a system using session types.
% , whose use are common in concurrency theory.
This allows us to introduce linear reasoning with few definitions.
With our objective focused on linearity, we forego channel delegation
and opt for a reduction approach rather than employing a transition
system semantics.  Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where restriction binds two names,n highlighting
their duality within the type system.  We have chosen a simple notion
of well-formedness, and do not  consider
deadlocks. While more sophisticated notions of well-formedness are
interesting, the proofs of these properties would distract from the
core linear aspects of the challenge.

\subsubsection{Scope extrusion.}
This is the challenge most closely related to the POPLMark challenge
since it concerns the properties of binders in restrictions and input.
It is not obvious what the limitations and advantages are in
leveraging the techniques used to solve the original challenges to
reason about scope extrusion.  For example, techniques like
higher-order abstract syntax or well-scoped de\ Bruijn indices lack
the concept of free variables and thus encoding this challenge with
them requires attention.

\subsubsection{Coinduction.}
Our coinduction challenge concerns strong barbed bisimulation and
congruence.\alb{we motivate why strong vs weak, but not why barbed}
Though weak barbed congruence is a more common behavioural equivalence
w.r.t.\ the \(\pi\)-calculus, here we prefer strong equivalences to avoid the
need to abstract over the number of internal transitions, thus
simplifying the theory.  We exclude delegation from the coinduction
challenge since it is orthogonal to our primary aim of exploring
coinductive proof techniques.

>>>>

% - Introduction
%   - POPLMark accelerated progress in mechanisation tools for PL
%     - POPLMark Reloaded tried to do the same for logical relations
%     - But they were deliberately limited in scope to not include process calculi
%   - We propose a new collection of benchmark problems for process calculi
%     - Why not concurrency in general? (after this point, only mention process calculi)
%     - Why is there a problem with process calculus mechanisations?
%     - Our experience says there is, and we will cite other experiences below
%   - We have noticed three key issues
%     - They are not covered by POPLMark
%     - Techniques are less settled than for binders
%     - Others also mention them as problems: cite and explain references briefly for each challenge topic
%   - Our goals are to determine the state of the art, develop best practices and tutorials, and identify improvements to proof assistant technology that would materially help the situation
\section{Introduction}
The POPLMark challenge~\cite{POPLMark} accelerated the development of proof assistants, libraries, and best practices for programming language mechanisation by establishing a common ground for systematically evaluating mechanisation techniques.
In doing so, it spearheaded a shift in the programming language research community towards publications that include mechanised proofs.
POPLMark Reloaded~\cite{POPLMarkReloaded} proposed a similar program for proofs involving logical relations.
These challenges, and therefore the developments they spurred, were deliberately limited in scope, but their authors noted the need for future benchmarks to tackle such areas as coinduction and linear environments.

@@

\todo{These paragraphs are still quite messy and unfocused. They should focus on the motivation, i.e.\ mention what others have perceived as problems.}
\paragraph{Linearity.}
In mechanised meta-theory, addressing linearity means choosing an appropriate representation of the linear context.
The crux of the matter is \emph{context splitting}, as showcased by the typing rule for parallel composition.

\todo{Insert related work focusing on noted issues with mechanising linearity.}

@@

\todo{Insert related work focusing on noted issues when mechanising scope extrusion.}

\cite{Maksimovic2015} does not have name restriction, thus dodging the challenge of scope extrusion.

@@

@@

\todo{Insert related work focusing on noted issues when mechanising coinduction.}

@@

% - Design of the benchmark
%   - We first list some general design remarks, then for each challenge
%   - General remarks
%     - We do not aim to be comprehensive (list other topics)
%     - However, we believe the best practices developed could also be of use in those topics
%     - We base our challenges on basic textbook theory with interesting proof techniques
%     - The challenges are independent (maybe note ``failure'' of POPLMark that everyone only did the first part)
%     - The challenges are as minimal as possible
%   - Linearity
%     - Type preservation for session types
%     - The main issue is context splitting
%     - Session types are pervasive in the field and simple to understand
%     - Binding two names makes duality obvious
%     - Brief overview of approaches and issues with them
%   - Scope extrusion
%     - Process calculi have several binders, but only restriction is different from ``normal'' ones
%     - The main issue is scope extrusion
%     - The pros and cons of established techniques for handling binders are different for process calculi, and it is unclear how exactly
%     - Brief overview of approaches and issues with them
%   - Coinduction
%     - Strong barbed bisimulation and congruence
%     - The main issue is the encoding of coinductive up-to techniques
%     - Strong equivalences to simplify (we don't have to abstract over internal transitions)
%     - Why barbed equivalences?
%     - No delegation to simplify
%     - Brief overview of approaches and issues with them
%   - Evaluation criteria
%     - We care not only about solutions, but about experiences and comparisons
%     - Mechanisation overhead
%     - Adequacy of the formal statements
%     - Cost of entry for tools and techniques
%     - Slight changes are OK if they are justified and adequate
\section{Design of the Benchmark}\label{sec:design-discussion}
\todo{Add references to support our claims and decisions in this whole section. Here is where we should mention existing approaches.}

@@

\subsubsection{Linearity.}
While a linear context is perhaps best seen as a multi-set, most proof assistants have better support for list processing.
The latter representation is intuitive, but may require establishing a large number of technical lemmata that are orthogonal to the mathematics of the problem under study, say proving type preservation for session types.
In this space, several designs are possible: one can label occurrences of resources to constrain their usage (\eg~\cite{CicconeP20}) or \alb{something --- ask Francisco} as in \cite{Castro2020}.
Some authors instead impose a multiset structure over lists (\eg~\cite{ChaudhuriLR19,Danielsson12}); still, the effort required to develop the infrastructure is significant.

\todo{Insert discussion/related work focusing on existing techniques and how our benchmark design makes it possible to compare them.}

For our challenge on linear reasoning we have chosen a type preservation theorem
for a system using session types.
% , whose use are common in concurrency theory.
This allows us to introduce linear reasoning with few definitions.
With our objective focused on linearity, we forego channel delegation
and opt for a reduction approach rather than employing a transition
system semantics.  Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where restriction binds two names,n highlighting
their duality within the type system.  We have chosen a simple notion
of well-formedness, and do not  consider
deadlocks. While more sophisticated notions of well-formedness are
interesting, the proofs of these properties would distract from the
core linear aspects of the challenge.

\subsubsection{Scope extrusion.}
This is the challenge most closely related to the POPLMark challenge
since it concerns the properties of binders in restrictions and input.

\todo{Insert discussion/related work focusing on existing techniques and how our benchmark design makes it possible to compare them.}

\subsubsection{Coinduction.}
Coinduction is widely supported in proof assistants.
Isabelle, Coq, and Abella seamlessly support coinductive definitions and reasoning.
Agda and Beluga support it through the convenient abstraction of co-patterns~\cite{Abel2013}.
Newer entrants to the field like the Lean proof assistant are working on adding a notion of coinduction to their system~\cite{Avigad2019,Keizer2023}.

\todo{Insert discussion/related work focusing on existing techniques and how our benchmark design makes it possible to compare them.}

Our coinduction challenge concerns strong barbed bisimulation and
congruence.\alb{we motivate why strong vs weak, but not why barbed}
Though weak barbed congruence is a more common behavioural equivalence
w.r.t.\ the \(\pi\)-calculus, here we prefer strong equivalences to avoid the
need to abstract over the number of internal transitions, thus
simplifying the theory.  We exclude delegation from the coinduction
challenge since it is orthogonal to our primary aim of exploring
coinductive proof techniques.
